Standardization of ceramic shape: A case study of Iron Age pottery from northeastern Taiwan

### 2020-09-08 analysis/paper/paper.Rmd
proof editing

---
title: "Standardization of ceramic shape: A case study from the Iron Age pottery from northeastern Taiwan"

author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The emergence of ceramic specialization in prehistoric societies is often linked to shifts in the complexity of social structures, because standardized ceramic production can reflect craft specialization and the presence of elite control. Previous work on identifying specialization relies on typological or linear metric analysis. Here we demonstrate how to investigate ceramic standardization by analyzing outlines of ceramic vessels. Outline analysis is useful because, unlike more commonly-used landmark analysis methods, it can effectively quantify shape differences for objects that lack distinctive measurement points needed for landmark analysis. We demonstrate this method using pottery from Kiwulan, a large multi-component Iron Age site (AD 1350-1850) in northeastern Taiwan. To measure ceramic specialization, we quantified pottery standardization by analyzing shape variables with reproducible geometric morphometric methods. We computed coefficients of variation (CVs) for shape coefficients obtained by elliptical Fourier analysis to test for shape standardization. We found significant differences in pottery shape and shape standardization that indicate changes in pottery production resulting from contact with mainland Han Chinese groups in northeastern Taiwan. Our case study, which includes an openly available research compendium of R code, represents an innovative application of outline-based methods in geometric morphometry to answer the anthropological questions of craft specialization. 
keywords: |
  Geometric Morphometrics; craft specialization; Iron age; culture contact; ceramic standardization
highlights: |
  1. Iron Age pottery shapes from Kiwulan, Taiwan, were analyzed using elliptical Fourier analysis
  2. Statistical tests for the equality of coefficients of variation of shape coefficients were applied
  3. Significant differences in pottery shape were identified between culture contact phases
  4. Pottery shapes show high standardization after contact with mainland Han Chinese groups

@@

For the degree of shape standardization, our CV tests on PC1 indicate a significant difference between the Chinese period and either pre-European or post-European periods. Analysis of morphological disparity, which measures the positioning of specimens relative to one another in the morphospace [@hopkins2017morphological], supports our finds of shape differences between Chinese contact and either pre-European (p = $`r pre_chi`$) or post-European periods (p= $`r post_chi`$) [@WangandMarwick2020], suggesting a more standardized shape after contact with the Chinese. In addition, we found a more homogeneous shape accompanied by a more standardized but also larger size in the Chinese period. Generally, people tend to make mistakes in hand-crafting as the size of an object increases, leading to higher variations in larger artifacts [@Eerkens2001]. However, we found the opposite for ceramics in the Chinese period when using body diameter as our proxy variable of size. This might hint an intentional behavior by Kiwulan potters to achieve a homogeneous form for the larger vessels. Mineral composition shows that the clay pastes are similar throughout three phases, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. We note that a small sample size in the Chinese period may lead to a more standardized shape. However, this effect can be reduced using CV statistics that scales variation to magnitude, allowing reliable comparison across uneven sample numbers, even for small sample size [@Eerkens2001]. Moreover, the MSLR test for equality of CVs enables a robust test between different sample numbers [@Krishnamoorthy2014].

@@

Additional insight into craft specialization at Kiwulan comes from the spatial pattern of ceramics, which provides information about potential production units and production areas [@Costin2001]. Figure \@ref(fig:kernel-plot) shows that the pottery samples have a widespread distribution with high densities of pottery at some units during the European presence. Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and a more extreme dispersed distribution after European presence. In contrast, the distribution of pottery is more similar to random distributions during the Chinese period. This is interesting because it contradicts our expectation that a clustered pattern will be observed with an increase in pottery standardization caused by the emergence of specialized groups [@Costin1995]. The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of the Indigenous population [@Chen2007; @Hsieh2009]. We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time. However, despite the small number of vessels during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time. As population across our three occupation phases declines, we see less clustered distributions of pottery, supporting an interpretation of intentional standardization rather than mechanical standardization. The spatial pattern shows that ceramics were mostly household-produced, and no specific facilities of production are evident [@Chen2007].

@@

We would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the pottery used in this study and providing the shape images. We thank Dr. Wen-Shan Chen in the Department of Geosciences, National Taiwan University for his invaluable guidance of petrographic analysis at his lab. We thank the Quaternary Research Center funding for supporting the organic residue analysis in this project. We thank Dr. Julian Sachs in the Department of Oceanography, University of Washington for his supports and suggestions for conducting organic geochemistry analysis of potsherds. We thank Dr. Matthew Wolhowe for his help in developing protocols for lipid extraction and his assistance with the GC-MS, GC-FID, and GC-C-IRMS analyses. This research used statistical consulting resources provided by the Center for Statistics and the Social Sciences, University of Washington. We also thank anonymous reviewers for their insightful comments that have greatly improved this manuscript.

>>>>

---
title: "Standardization of ceramic shape: A case study of Iron Age pottery from northeastern Taiwan"

author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The emergence of ceramic specialization in past societies is often linked to shifts in the complexity of social structures, because standardized ceramic production can reflect craft specialization and the presence of elite control. Previous work on identifying specialization relies on typological or linear metric analysis. Here we demonstrate how to investigate ceramic standardization by analyzing outlines of ceramic vessels. Outline analysis is useful because, unlike more commonly-used landmark analysis methods, it can effectively quantify shape differences for objects that lack distinctive measurement points needed for landmark analysis. We demonstrate this method using pottery from Kiwulan, a large multi-component Iron Age site (AD 1350-1850) in northeastern Taiwan. To measure ceramic specialization, we quantified pottery standardization by analyzing shape variables with reproducible geometric morphometric methods. We computed coefficients of variation (CVs) for shape coefficients obtained by elliptical Fourier analysis to test for shape standardization. We found significant differences in pottery shape and shape standardization that indicate changes in pottery production resulting from contact with mainland Han Chinese groups in northeastern Taiwan. Our case study, which includes an openly available research compendium of R code, represents an innovative application of outline-based methods in geometric morphometry to answer the anthropological questions of craft specialization. 
keywords: |
  Geometric Morphometrics; craft specialization; Iron age; culture contact; ceramic standardization; Taiwan
highlights: |
  1. Iron Age pottery shapes from Kiwulan, Taiwan, were analyzed using elliptical Fourier analysis
  2. Statistical tests for the equality of coefficients of variation of shape coefficients were applied
  3. Significant differences in pottery shape were identified between culture contact phases
  4. Pottery shapes show high standardization after contact with mainland Han Chinese groups

@@

For the degree of shape standardization, our CV tests on PC1 indicate a significant difference between the Chinese period and either pre-European or post-European periods. Analysis of morphological disparity, which measures the positioning of specimens relative to one another in the morphospace [@hopkins2017morphological], supports our finds of shape differences between Chinese contact and either pre-European (p = $`r pre_chi`$) or post-European periods (p= $`r post_chi`$) [@WangandMarwick2020], suggesting a more standardized shape after contact with the Chinese. In addition, we found a more homogeneous shape accompanied by a more standardized but also larger size in the Chinese period. Generally, people tend to make mistakes in hand-crafting as the size of an object increases, leading to higher variations in larger artifacts [@Eerkens2001]. However, we found the opposite for ceramics in the Chinese period when using body diameter as our proxy variable of size. This might hint an intentional behavior by Kiwulan potters to achieve a homogeneous form for the larger vessels. Mineral composition shows that the clay pastes are similar throughout three phases, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. We note that a small sample size in the Chinese period may lead to a more standardized shape. However, this effect can be reduced using CV statistics that scales variation to magnitude, allowing reliable comparison across uneven sample numbers, even for small sample size [@Eerkens2001]. Moreover, the MSLR test for equality of CVs enables a robust test between different sample sizes [@Krishnamoorthy2014].

@@

Additional insight into craft specialization at Kiwulan comes from the spatial pattern of ceramics, which provides information about potential production units and production areas [@Costin2001]. Figure \@ref(fig:kernel-plot) shows that the pottery samples have a widespread distribution with high densities of pottery at some units during the European presence. Hypothesis testing on spatial randomness (Figure \@ref(fig:Monte-Carlo-spatial-pattern)) indicates a non-randomly dispersed distribution before European contact and a more extreme dispersed distribution after European presence. In contrast, the distribution of pottery is more similar to random distributions during the Chinese period. This is interesting because it contradicts our expectation that a clustered pattern will be observed with an increase in pottery standardization caused by the emergence of specialized groups [@Costin1995]. The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of the Indigenous population [@Chen2007; @Hsieh2009]. We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time. However, despite the small number of vessels during the Chinese contact period, Figure \@ref(fig:Monte-Carlo-spatial-pattern) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time. As population across our three occupation phases declines, we see less clustered distributions of pottery, supporting an interpretation of intentional standardization rather than mechanical standardization. The spatial pattern shows that ceramics were mostly household-produced, and no specific facilities of production are evident [@Chen2007] (Figure \@ref(fig:kernel-plot)).

@@

We would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the pottery used in this study and providing the shape images. We thank Wen-Shan Chen in the Department of Geosciences, National Taiwan University for his invaluable guidance of petrographic analysis at his lab. We thank the Quaternary Research Center funding for supporting the organic residue analysis in this project. We thank Julian Sachs in the Department of Oceanography, University of Washington for his supports and suggestions for conducting organic geochemistry analysis of potsherds. We thank Matthew Wolhowe for his help in developing protocols for lipid extraction and his assistance with the GC-MS, GC-FID, and GC-C-IRMS analyses. This research used statistical consulting resources provided by the Center for Statistics and the Social Sciences, University of Washington. We also thank Ben Fitzhugh, Peter Lape, Robert Z. Selden Jr. and the anonymous reviewers for their insightful comments that have greatly improved this manuscript.
